---
title: "Assignment Week 5"
author: "Ruqia Ali Hassan"
date: "2025-03-25"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Loading Libraries

First, all of the necessary libraries for genome analysis are loaded.

```{r}
suppressPackageStartupMessages({
  library(AnnotationHub)
  library(ensembldb)
  library(GenomicRanges)
  library(epiwraps)
  library(rtracklayer)
  library(ggplot2)
})
```

# Bivalent domains in between H3K27me3 and H3K4me3 in mouse embryonic stem cells (mESC) and erythroblasts

## Download files

The species being analyzed is the mouse. For each marker (H3K27me3 and H3K4me3), we download data for the embryonic stem cell line (mESC) and a differentiated cell line (erythroblasts).

```{r}
options(timeout=1800)
dir.create("peaks")
#Embryonic cells

#H3K27me3 
download.file("https://www.encodeproject.org/files/ENCFF008XKX/@@download/ENCFF008XKX.bed.gz", "peaks/H3K27me3_mESC.bed.gz")

#H3K4me3
download.file("https://www.encodeproject.org/files/ENCFF974BMC/@@download/ENCFF974BMC.bed.gz","peaks/H3K4me3_mESC.bed.gz")

#Differentiated cells

#H3K27me3 
download.file("https://www.encodeproject.org/files/ENCFF728FZO/@@download/ENCFF728FZO.bed.gz", "peaks/H3K27me3_diff.bed.gz")

#H3K4me3
download.file("https://www.encodeproject.org/files/ENCFF579NUO/@@download/ENCFF579NUO.bed.gz","peaks/H3K4me3_diff.bed.gz")


```

### Identify bivalent domains

A peak file is created to read all four files.

```{r}
#list the files
peakfiles <- list.files("peaks", full.names = TRUE)

#give them meaningful names
names(peakfiles) <- gsub("\\.bed.gz","",basename(peakfiles))

#read the peaks:
peaks <- lapply(peakfiles, FUN=rtracklayer::import, format="narrowPeak")
#print(peaks)

#saveRDS(peaks, file="peaksBug.rds") 
#sapply(peaks, class)
#lengths(peaks) 
```

```{r}
# Bivalent fraction calculation for embryonic cells (not needed her but done to better understand the lapply function)

#Direct approache
#bivalent_fraction <- sum(overlapsAny(peaks$H3K27me3, peaks$H3K4me3)) / length(peaks$H3K27me3) *100
#print(bivalent_fraction)

#Lapply approache
#bivalent_fraction_lapply <- lapply(peaks, FUN=function(x) sum(overlapsAny(peaks$H3K27me3, x)) / length(peaks$H3K27me3) *100)
#print(bivalent_fraction_lapply) 

#The function(x) in lapply() iterates through all peak files, including self-comparisons (e.g., H3K27me3 vs H3K27me3), which calculates self-overlap. In contrast, the direct approach overlap_direct <- sum(overlapsAny(peaks$H3K27me3.gz, peaks$H3K4me3.gz))/length(peaks$H3K27me3.gz) specifically measures overlap between different peak sets (H3K27me3 and H3K4me3) without including self-overlaps.

#can be done for the differentiated cell, but was skipped as fraction was not asked for. 
```

#### Bivalent domains in embryonic stem cells

Using findOverlaps(), we identify the peaks that overlap between the two markers in mESCs. I used two different approaches to determine the bivalent domains, with the second being more precise.

```{r}
#Find the bivalent domains
mESC_overlaps <- findOverlaps(peaks$H3K4me3_mESC, peaks$H3K27me3_mESC)

# H3K27me3 peaks that overlap with H3K4me3 (full peaks)
bivalent_domains_mESC <- peaks$H3K27me3_mESC[subjectHits(mESC_overlaps)]
print(bivalent_domains_mESC)
#print(width(bivalent_domains_mESC))  # Widths of full H3K27me3 peaks

#EXACT overlapping domains
exact_bivalent_domains_mESC <- pintersect(
peaks$H3K4me3_mESC[queryHits(mESC_overlaps)], 
peaks$H3K27me3_mESC[subjectHits(mESC_overlaps)])
print(exact_bivalent_domains_mESC)
#print(width(exact_bivalent_domains_mESC))  # Widths of precise overlaps

#total_bivalent_domains_mESC <- length(exact_bivalent_domains_mESC)
#print(total_bivalent_domains_mESC)
```

regionOverlaps() and regionUpset() are used to visualize the overlaps within the two markers in mESCs.

```{r}
#Visualizing the overlaps
regionOverlaps(peaks)
regionUpset(peaks, nsets=length(peaks))
```

### How many of the mESC bivalent domains are, in this differentiated cell type, overlapping either mark or their combination (in this differentiated cell type)?

We find the overlaps of mESC bivalent domains with the markers in erythroblasts using findOverlaps(). To avoid double counting, setdiff() is used. The total unique overlaps are then calculated.

```{r}
# Overlaps with erythroblast histon markers
overlap_H3K27me3 <- findOverlaps(exact_bivalent_domains_mESC, peaks$H3K27me3_diff) 
overlap_H3K4me3  <- findOverlaps(exact_bivalent_domains_mESC, peaks$H3K4me3_diff)
#findOverlaps() returns all the overlaps including the domains that overlap with both H3K27me3 and H3K4me3

# Get UNIQUE (remove doubles) mESC domains overlapping each mark/both
H3K27me3_domains <- unique(queryHits(overlap_H3K27me3))  
H3K4me3_domains <- unique(queryHits(overlap_H3K4me3))   
both_marker_domains <- intersect(H3K27me3_domains, H3K4me3_domains)  

# Total unique domains overlapping either mark or both (union of the two sets)
total_unique_overlap <- length(union(H3K27me3_domains, H3K4me3_domains))

# Counts
count_H3K27me3_only <- length(setdiff(H3K27me3_domains, H3K4me3_domains)) #setdiff returns the elements that are present in one set but not in another 
count_H3K4me3_only  <- length(setdiff(H3K4me3_domains, H3K27me3_domains))
count_both <- length(both_marker_domains)

# Print results
cat("Total mESC bivalent domains overlapping either/both marks in erythroblasts:", total_unique_overlap, "\n")
cat(" - Overlapping H3K27me3 only:", count_H3K27me3_only, "\n")
cat(" - Overlapping H3K4me3 only:", count_H3K4me3_only, "\n")
cat(" - Overlapping both marks:", count_both, "\n")
```

The data reveals that 4192 bivalent domains in mouse embryonic stem cells (mESCs) overlap with histone modifications in erythroblasts, with 2822 overlapping H3K27me3, 101 overlapping H3K4me3, and 1269 overlapping both.

```{r cars}
summary(cars)
```
